---
title: "writeup"
author: "Jingyuan Wu"
date: "2021/9/24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Birthday Problem

*In a class of N individuals, what is the probability that at least two students will share a birthday?*

## Introduction

This blog is going to explain how to solve the birthday problem with both simulation and analytic tools. A plot that shows both solutions for possible class sizes from 1 to 80 will be listed at last.

## Assumptions

Leap year is ignored.

Birthdays are equally likely to fall on any day of the year.

## Limitations

Everyone’s birthday is independent.

365 days are in a 1-365 order.

## Solution1: Analytics

Assume: 

There are n students in a class.

P(A) = {the probability that at least two students will share a birthday}

The complement of A: P(A') = {the probability that every students has a distinct birthday}

P(A) = 1 - P(A')

After ignoring the leap year, there are 365 days in a year. Each student can select any one of these 365 days. As a result, for n students, the sample space will be: 365^n.

According to the assumption, if every students has a distinct birthday, student1 will have 365 days to choose, student2 will have 364 days to choose, student3 will have 363 days to choose, ... The number of combinations that n students select n distinct days from 365 days will be: 365!/(365-n)!, which is the total number of sequence without replacement.

The probability that every student has a distinct birthday: 365!/(365-n)!/(365^n)

P(A') = [365!/(365-n)!]/(365^n)

P(A) = 1-P(A') = 1-[365!/(365-n)!]/(365^n) = 1-n!(365, n)/(365^n)

```{r}
library(ggplot2)
library(tidyverse)

n <- 80
p <- rep(NA, n)
for (i in 1:n){
  p[i] <- 1-choose(365, i) * factorial(i)/(365^i)
  p[1:i]
}

bdp_a <- data.frame(class_size=1:n, p)
#bdp_a
```

The code above is using a for loop to calculate the probability that at least two students will share a birthday for the class sizes from 1 to 80 with the analytic solution.

## Solution2: Simulation

```{r}
z=0
pd <- vector()
q <- vector()
times <- 1000

for (x in 1:n){
  for (y in 1:times){
    bdp_s <- data.frame(class_size=1:x, birthday = round(runif(x,min=1,max=365), 0))
    bdp_s1 <- bdp_s %>% group_by(birthday) %>% summarise(sum_bd=n())
    pd[y]=ifelse(nrow(bdp_s1)<x, z+1, z+0)
      }
  q[x]<-mean(pd)
}
#q

bdp_s2 <- data.frame(class_size=1:n, q)

g2<- bdp_s2 %>% ggplot(aes(x=class_size, y=q)) +
            geom_point(aes(col="Simulation solution")) +
   labs(x="Size of group", y="Probability of at least one shared birthday") +
   theme_bw() +
   theme(panel.grid.major=element_blank(), panel.grid.minor = element_blank())
```

The code above is using a for loop to calculate the probability that at least two students will share a birthday for the class sizes from 1 to 80 with the simulation solution. The simulation is using another for loop to simulate 1000 times.

## Two Solutions Shown in One Plot

```{r}
g2 +
   geom_point(aes(class_size, p, col="Analytic solution")) +
   geom_hline(yintercept=c(0,1), linetype="longdash", color="gray")
```

After plotting two solutions in one graph, they shows similar trends. The greater the class size is, the higher the probability that at least two students will share a birthday is, which finally is even close to 1. However, the points of simulation solution seem more deviated. They scatter around the points of analytic solution.

## Pros and Cons of both Approaches

Analytics

Pros: For this problem, analytics seems concise, more accurate and time-saving.

Cons: Analytics needs basic knowledge of combinatorics and binomial distribution. If the calculation of the question is too complicated due to the sample size/the number of combinations, it will be hard for analytics to operate.


Simulation

Pros: Simulation can be easily replicated and the conditions of simulation can be varied.

Cons:  For this problem, simulations seems time-consuming and not precise enough. Simulations are always based on and limited to theories/hypotheses/conditions and can never take place of real tests. Deviations can not be avoided in simulations due to the algorithms of computers. Since the more simulations, the closer the results are to the true values, sometimes, simulations may cause extra time costs for higher accuracy.
