---
title: "writeup"
author: "Jingyuan Wu"
date: "2021/9/26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## How often does the better team win the World Series?

### Introduction

World Series is a best-of-7 match-up between the champions of the American and National Leagues of Major League Baseball. This blog is going to explain the probability calculation problems related to World Series.

### Setup

Suppose that the Braves and the Yankees are teams competing in World Series.

Suppose that in any given game, the probability that the Braves wins is PB and the probability that the Yankees wins is PY = 1 − PB.

### Assumptions

The result of each game is independent.

The result of each game is either a win or a loss, no tie.

When it comes to best-of-X series, X should be odd.

The team that wins World Series must win the last game.

### Q1 What is the probability that the Braves win the World Series given that PB = 0.55?

#### Setup for Q1
The probability that the Braves wins in any given game is 0.55.

Since it is a best-of-7 match-up, the Braves must win 4 games to win the series and could end the series at the 4th or 5th or 6th or 7th game. For each combination of wins and losses, the Braves must win the last game.

If the series ends in the 4th game, the Braves must win all 4 games and loose 0 game.

If the series ends in the 5th game, the Braves must win the last game and any 3 other games before, and can loose any 1 game before the last game.

If the series ends in the 6th game, the Braves must win the last game and any 3 other games before, and can loose any 2 games before the last game.

If the series ends in the 7th game, the Braves must win the last game and any 3 other games before, and can loose any 3 game before the last game.

The combinations and their probabilities are shown below.

| End_n | losses |     probability    |
|:-----:|:------:|:------------------:|
|   4   |   0    |  dnbinom(0,4,.55)  |
|   5   |   1    |  dnbinom(1,4,.55)  |
|   6   |   2    |  dnbinom(2,4,.55)  |
|   7   |   3    |  dnbinom(3,4,.55)  |

The probability that the Braves wins the World Series given that PB = 0.55: $$dnbinom(0,4,.55)+dnbinom(1,4,.55)+dnbinom(2,4,.55)+dnbinom(3,4,.55)=pnbinom(3,4,.55)$$

```{r}
library(ggplot2)
library(tidyverse)
pnbinom(3,4,.55)
```

The probability that the Braves wins the World Series given that PB = 0.55 is 0.61.

### Q2 What is the probability that the Braves win the World Series given that PB = x? This will be a figure (see below) with PB on the x-axis and P(Braves win World Series) on the y-axis.

#### Setup for Q2
The probability that the Braves wins in any given game is x.
x: [0.5, 1], break=0.01.

As it is still a best-of-7 match-up, the calculation formula remains unchanged, except that the probability that the Braves wins in any given game turns to x.

```{r}
x=seq(0.5, 1, 0.01)
y=pnbinom(3,4,x)
data <- data.frame(x,y)
data %>% ggplot(aes(x, y)) +
   geom_point() +
   #ylim(0,1) +
   scale_y_continuous(limit=c(0,1),breaks = seq(0, 1, .2)) +
   labs(x="Probability of the Braves Winning a Head-to-Head Matchup", y="Pr(Win World Series)", title="Probability of Winning the World Series") +
   theme_bw()  +
   theme(plot.title = element_text(hjust = 0.5), panel.grid.major=element_blank(), panel.grid.minor = element_blank())
```

The graph above shows the relationship between `Probability of the Braves Winning a Head-to-Head Matchup` and `Pr(Win World Series)`. The higher the `Probability of the Braves Winning a Head-to-Head Matchup` is, the higher `Pr(Win World Series)` is.

### Q3 Suppose one could change the World Series to be best-of-9 or some other best-of-X series. What is the shortest series length so that P(Braves win World Series|PB = .55) ≥ 0.8

#### Setup for Q3
The probability that the Braves wins in any given game is 0.55.

This time World Series will be a best-of-X series. X is changing from 1 to 99 and breaking in 2. A for loop is set for X to find the shortest series length. While P(Braves win World Series|PB = .55) ≥ 0.8, the for loop breaks and outputs the shortest series length.

The calculation formula remains unchanged. `ceiling()` function is used for returning the smallest integer that is greater than or equal to the value passed to it as argument, which is in order to find the `losses at most` of winning a best-of-X series. 

```{r}
X=99 #X should be odd.
Z=0
Y <- vector()
for (i in seq(1,X,2)){
  Y[i]=pnbinom(ceiling(i/2)-1, ceiling(i/2), .55)
  if (Y[i]>=0.8){
    Z=i
    break
  }
}
Z
```

The shortest series length given P(Braves win World Series|PB = .55) ≥ 0.8 is 71.

### Q4 What is the shortest series length so that P(Braves win World Series|PB = x) ≥ 0.8? This will be a figure (see below) with PB on the x-axis and series length is the y-axis.

#### Setup for Q4
The probability that the Braves wins in any given game is m.
m: [0.5, 1], break=0.01.

This time World Series will be a best-of-q series. q should be odd.

The calculation formula remains unchanged, except that the probability that the Braves wins in any given game turns to `m[j]`, the Braves at most loses `[ceiling(q[i]/2)-1]` games.

`t[j]` is the shortest length of series that Braves wins World Series given `m[j]`.

Double for loop is set for iterations of `Probability of the Braves Winning a Head-to-Head Matchup` and `Series Length`.

```{r}
m=seq(0.5, 1, 0.01)
q=seq(1, 9999, 2) #q should be odd.
t=rep(NA, length(m))
s=rep(NA, length(q))
for (j in 1:length(m)){
  for (i in 1:length(q)){
    r=pnbinom(ceiling(q[i]/2)-1, ceiling(q[i]/2), m[j])
  if(r>=0.8){
    s=q[i]
    break
  }
}
t[j]=s
}
t

data1=data.frame(m, t)
#sum(is.na(data1))
max <- data1 %>% na.omit() %>% filter(t==max(t))
#max
data1 %>% na.omit() %>% ggplot(aes(m, t)) +
   geom_point() +
   labs(x="Probability of the Braves Winning a Head-to-Head Matchup", y="Series Length", title="Shortest Series so that P(Wins WS given p)>=0.8") +
   theme_bw() +
   theme(plot.title = element_text(hjust = 0.5), panel.grid.major=element_blank(), panel.grid.minor = element_blank()) +
   geom_text(data=max, label=max$t, vjust=1.5, size=3)
```

The graph above shows the relationship between `Probability of the Braves Winning a Head-to-Head Matchup` and `Series Length`. The higher the `Probability of the Braves Winning a Head-to-Head Matchup` is, the shorter `Series Length` is.

### Q5 Calculate P(PB = 0.55|Braves win World Series in 7 games) under the assumption that either PB = 0.55 or PB = 0.45. Explain your solution.

#### Setup for Q5
A = {PB = 0.55}

B = {Braves win World Series in 7 games}

C = {PB = 0.45}

According to `Bayes Rules`, $$P(A|B) = \frac{P(B|A)P(A)}{P(B)}$$

As there are two possible PB listed in Q5, it is easy to find that: $$P(A) = P(C) = 0.5$$

According to the `Law of Total Probability`:
$$P(B) = P(B|A) + P(B|C)$$

$$P(B) = dnbinom(3,4,.55) *.5 + dnbinom(3,4,.45)] * .5$$

When calculating the probability of the Braves wins World Series in 7 games given the probability that the Braves wins in any given game is 0.55, it is obvious that the Braves wins the last game of the best-of-7 match-up and loses any 3 of the previous games. As a result, $$P(B|A) = dnbinom(3,4,.55)$$

$$P(A|B) = \frac{P(B|A)P(A)}{P(B)} = \frac{dnbinom(3,4,.55)*0.5}{dnbinom(3,4,.55) *.5 + dnbinom(3,4,.45)*0.5}$$

```{r}
pb<-dnbinom(3,4,.55)*.5+dnbinom(3,4,.45)*.5 #P(Braves win World Series in 7 games)
pba=dnbinom(3,4,.55) #P(Braves win World Series in 7 games|P_B=0.55)
pa=.5 #P(P_B=0.55)
pab=pba*pa/pb
pab #Answer
```

P(PB = 0.55|Braves win World Series in 7 games) under the assumption that either PB = 0.55 or PB = 0.45 is 0.55.
